Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-07-29 writeups/paper.tex
some first edits

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

% cp ../sims/agg.png plots
% cp ../sims/2d_3_final/snapshots_lin_0_masked/f_amps.png plots/lin_amps.png
% cp ../sims/2d_3_final/snapshots_lin_0_masked/fluxes.png plots/lin_fluxes.png
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps.png plots/nl_f_amps.png
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/fluxes.png plots/nl_fluxes.png
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_refl.png plots/nl_f_refl.png
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_ri.png plots/nl_f_ri.png
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/front.png plots/nl_front.png

@@

We consider a two-dimensional incompressible, isothermal fluid, representative
of degenerate matter in WDs. We furthermore neglect temperature variations and
assume a barotropic equation of state $P(\rho, T) = P(\rho)$ as a first
approximation. As we are interested in dynamics far from the center of the WD,
we approximate the gravitational field as uniform. We model the background
density stratification as $\rho_0(x, z) = \rho_0(z) = \rho_0(z=0) e^{-z/H}$ for
some reference density $\rho_0(z=0) = \at{\rho_0(z)}_{z = 0}$.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\vec{u}_1}{t} + \frac{\vec{\nabla}p}{\rho} + g\hat{z} &=
            0.\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\vec{u}_1 \cdot \vec{\nabla}}$ is the Lagrangian or
material derivative, and $\vec{u}_1, \rho, p$ denote the velocity field, density
and pressure respectively. We denote $-g\hat{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 = \rho_0 gH$. A shear flow of
form $\vec{u}_1 = u_{1x}(x, z, t)\hat{x} = u_{1x}(z)\hat{x}$ is permitted in hydrostatic
equilibrium, but we will assume no background shear flow. Physically, this
assumption corresponds to a non-rotating star, or going to the corotating frame
of a rigidly rotating star.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\rho_0}$. This both identically enforces $\rho > 0$ and avoids
numerical issues in the $\frac{\vec{\nabla}p}{\rho}$ term if $\rho$ is small. We
also define reduced pressure $P = \frac{p}{\rho}$. Then, we may rewrite the
second two equations of \autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_{1z} \pd{\ln \rho_0}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\vec{u}_1}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\Upsilon_0 = 1, P_0 = gH$.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \vec{u}_1 \cdot \vec{\nabla}$,
we may linearize \autoref{se:nl_upsilon}. We ignore the advective components of the
material derivative and consider small deviations $\Upsilon_1 = \Upsilon -
\Upsilon_0, P_1 = P - P_0$ about hydrostatic equilibrium:
\begin{subequations}\label{se:lin_homo}
    \begin{align}
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
        \pd{\vec{u}_1}{t} + \vec{\nabla}P_1
            + \p{1 + \Upsilon_1} g\hat{z}
            &= 0.
    \end{align}
\end{subequations}

This can be solved to obtain solutions up to undetermined amplitude $A$ the
well-known result \citep{drazin,sutherland0}:
\begin{equation}
    u_{1z}\p{x, z, t} = Ae^{z/2H}e^{i\p{k_{1x}x - k_{1z}z - \omega_1 t}},
        \label{eq:k0z_sign}
\end{equation}
where $\omega_1$ satisfies dispersion relation
\begin{equation}
    \omega_1^2 = \frac{N^2k_{1x}^2}{k_{1x}^2 + k_{1z}^2 + \frac{1}{4H^2}},
        \label{eq:disp_rel}
\end{equation}
where
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}} = \frac{g}{H},
\end{equation}
the Brunt-V\"ais\"al\"a frequency is constant. Our sign convention is such that
a wave with $k_{1x}, k_{1z} > 0$ has positive vertical group velocity (see
\autoref{eq:vg}).

In the weak stratification limit $k_{1z}H \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows with $z$ as $e^{z/2H}$. Thus, the
        linear approximation is always violated for sufficiently large $z$.

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \vec{c}_{ph} &=
                \p{k_{1x}\hat{x} - k_{1z}\hat{z}}\frac{\omega_1}
                {k_{1x}^2 + k_{1z}^2 + \frac{1}{4H^2}},\\
            \vec{c}_{g} &= N\frac{\p{k_{1z}^2 + \frac{1}{4H^2}}\hat{x}
                + \p{k_{1x}k_{1z}\hat{z}}}
                {\p{k_{1x}^2 + k_{1z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\vec{c}_{ph} \cdot \vec{c}_g = +
        \mathcal{O}\p{\frac{1}{\p{k_{1z}H}^{2}}} \approx 0$, the usual result
        (e.g.\  \citep{drazin,sutherland1}). Note again that $c_{gz} > 0$ for
        $k_{1x}, k_{1z} > 0$, a consequence of our sign choice in
        \autoref{eq:k0z_sign}.

    \item The time-averaged total $x$-momentum flux carried in the $\hat{z}$
        direction can be computed. Since the linear solution is separable as
        $f(x, z, t) = f(z)e^{i(k_{1x}x - \omega_1 t)}$, $x$ averaging and time
        averaging are equivalent, so we may write
        \begin{align}
            S &\equiv \ev{\rho u_{1x} u_{1z}}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{1x} u_{1z}\;\mathrm{d}x,
                    \label{eq:S_def},\\
                &\approx -\frac{A^2}{2}\rho_0(z=0)\frac{k_{1z}}{k_{1x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_{1z}\pd{\ln \rho_0}{z}
        = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{1x}x - \omega_1 t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{1z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the
given $k_{1x}, \omega_1(k_{1x}, k_{1z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two homogeneous solutions above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{1z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z = k_{1z}$), and we obtain
\begin{align}
    u_{z}&(x, z, t) ={} \frac{Fgk_{1x}^2}{\omega_1^2}
        \frac{1}{2ik_{1z}}\frac{e^{\frac{-(k_{1z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{1x}x + k_{1z}(z - z_0) - \omega_1 t
            + \frac{1}{2k_{1z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{1x}x - k_{1z}(z - z_0) - \omega_1 t
            + \frac{1}{2k_{1z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ solution models an upwards-propagating IGW wavetrain inbound on
the simulation domain from below. The solution for $u_{1x}(x, z, t)$ can be
obtained analytically from incompressibility \autoref{eq:nl_incomp}.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u}_1 \cdot \vec{\nabla}$. However, since $\abs{\vec{u}_1} \propto
e^{z/2H}$, any nonzero amplitude IGW will eventually violate the linearity
criterion. This regime is relevant in practice since the IGW may grow to
nonlinear amplitudes before it reaches the surface of the WD
\citep{fullerI,fullerII}. We can coarsely estimate the height at which this
happens by when the Lagrangian displacement $\vec{\xi} =
\frac{\vec{u}_1}{-i\omega}$ of a fluid parcel satisfies
\begin{equation}
    \vec{\xi} \cdot \vec{k}_1 \gtrsim 1.\label{eq:nl}
\end{equation}

Another useful result quantifies the wave-induced mean flow
\citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{1x}} \equiv \bar{U} = \frac{1}{L_x}
        \int\limits u_{1x}\;\mathrm{d}x = \frac{\ev{u_{1x}u_{1z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $S$ (\autoref{eq:S_def}) into a medium at
rest. Solving for $\bar{U} = c_{ph, x}$ the critical layer condition (see
\autoref{ss:crit_layer} immediately below), we recover
$\frac{u_{1x}k_{1x}}{\omega} = \xi_x k_{1x} \sim 1$, roughly equivalent to the
nonlinearity criterion \autoref{eq:nl}. This condition predicts where the wave's
self-acceleration alone is sufficient to induce critical layer absorption.

@@

A purely horizontal shear flow $\bar{U}(z) \hat{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\bar{U}(z)\partial_x$. For a critical value $\omega_1 - \bar{U}_c k_{1x} = 0$,
the frequency of the linear wave in the fluid's frame of reference vanishes and
critical behavior is observed. In a linear \citep{booker_bretherton} or viscous
\citep{hazel} theory, the incident wave has amplitude reflection and
transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\bar{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\bar{U}(z_c) = \frac{\omega_1}{k_{1x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the shear flow cannot exceed $\bar{U}_c$ the horizontal
phase velocity of the incident wave, the critical layer must instead propagate
downwards to accommodate the incident momentum. The total horizontal momentum of
the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits_0^{L_z} \rho(z) \bar{U}(z, t)\;\mathrm{d}z
        + \Delta S = 0.
\end{equation}
We define $\Delta S$ to be the change in flux across the critical layer, or
equivalently the negative of the absorbed flux. Treating $\bar{U}(z > z_c) =
\bar{U}_c, \bar{U}(z < z_c) = 0$ gives us exactly
\begin{equation}
    \rho(z_c) \bar{U}_c\pd{z_c}{t} = \Delta S(z_c(t)).\label{eq:zc_anal}
\end{equation}
$\Delta S(z_c)$ denotes the change in $S$ across $z_c$. For constant $\Delta S$
in time and $\rho \approx \rho_0$, this has analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\rho_0(z = 0)c_{ph, x}}{-\Delta S}
        .\label{eq:zc_sol}
\end{equation}
Note that constant $-\Delta S$ corresponds physically to the picture where
constant $S$ carried by the IGWs, given by \autoref{eq:S_lin}, is fully absorbed
at the critical layer, in which case $\Delta S = -S$.

@@

We nondimensionalize by taking $H = N = \rho_0(z=0) = 1$ in
\autoref{se:nl_orig}.

Denote $L_x, L_z$ to be the physical dimensions of the simulation domain. Our
simulation domain is smaller than our full physical domain, so we adopt
reflection-suppressing boundary conditions. We use periodic boundary conditions
in the $x$ direction and damping layers (described in \autoref{ss:damping}) in
the $z$ direction to damp perturbations exiting the top and bottom of the
domain. We use a Fourier basis in the $x$ direction and either a Chebyshev or
Fourier basis in the $z$ direction; our results were robust to either set of
basis functions. We varied the number of $x, z$ modes (denoted $N_x, N_z$
respectively), and used $3/2$ dealiasing \citep{boyd}.

@@

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from advective Courant-Friedrichs-Lewy (CFL)
time. Specifically, we use $\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_{1z})$,
where the minimum is taken over every grid point in the domain and $\Delta
x,\Delta z$ are the grid spacings in the $x$ and $z$ directions respectively.

The physics of our simulation is fixed by four parameters: $k_{1x}, \omega_1, F,
\nu$. We describe our choices for these parameters below:
\begin{itemize}
    \item $k_{1x}$: Astrophysical IGWs in stars are generally excited by the $l
        = 2$ component of the tidal potential, for which $k_{\perp} \sim
        \frac{1}{R}$ where $R$ is the radius of the star. To best emulate this,
        we use $k_{1x} = \frac{2\pi}{L_x}$ the smallest permitted wavenumber
        permitted by periodic boundary conditions.

    \item $\omega_1$: We choose $\omega_1$ by evaluating dispersion relation
        $\omega_1\p{k_{1x}, k_{1z}}$ for some desired $k_{1z}$.

        Astrophysical IGWs also generally satisfy $\omega \ll N$, or
        equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since $H \lesssim
        R$, we aim to study $k_{1z}H \gg 1$. However, to satisfy this well is
        computationally expensive: $L_z \gg H$ is required to give waves ample
        room to grow within the simulation domain, but we want $k_{1z}$ to be
        sufficiently separated from the grid spacing $\sim L_z / N_z$ that
        nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{1z}H
        \gg 1$; we choose $k_{1z}H = 1$.

>>>>

    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\langle#1\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\boldsymbol{\mathbf{#1}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}

@@

% cp ../sims/agg.png plots;
% cp ../sims/2d_3_final/snapshots_lin_0_masked/f_amps.png plots/lin_amps.png;
% cp ../sims/2d_3_final/snapshots_lin_0_masked/fluxes.png plots/lin_fluxes.png;
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps.png plots/nl_f_amps.png;
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/fluxes.png plots/nl_fluxes.png;
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_refl.png plots/nl_f_refl.png;
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_ri.png plots/nl_f_ri.png;
% cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/front.png plots/nl_front.png;

@@

We consider a two-dimensional incompressible, isothermal fluid, representative
of degenerate matter in WDs. We furthermore neglect temperature variations and
assume a barotropic equation of state $P(\rho, T) = P(\rho)$ as a first
approximation. As we are interested in dynamics far from the center of the WD,
we approximate the gravitational field as uniform. We model the background
density stratification as $\overline{\rho}(x, z) = \overline{\rho}_0 e^{-z/H}$
for some reference density $\overline{\rho}_0$ (we generally notate background
quantities with overbars and perturbation quantities with primes).

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\bm{z} &=
            0\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\bm{u} \cdot \bm{\nabla}}$ is the Lagrangian or
material derivative, and $\bm{u}, \rho, P$ denote the velocity field, density
and pressure respectively. We denote $-g\bm{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\bm{\nabla}\overline{P} = -\overline{\rho} g\bm{z}$ and so $\overline{P} =
\overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\bm{x}$ is permitted in hydrostatic equilibrium, but we will
assume no background flow. Physically, this assumption corresponds to a
non-rotating star, or going to the corotating frame of a rigidly rotating star.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$. This both identically enforces $\rho > 0$ and
avoids numerical issues in the $\frac{\bm{\nabla}P}{\rho}$ term if $\rho$ is
small. We also define reduced pressure $\varpi = \frac{P}{\rho}$. Then, we may
rewrite the second two equations of \autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u'_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi
            + \varpi\bm{\nabla}\Upsilon + g\bm{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\overline{\Upsilon} = 1, \overline{\varpi} = gH$.

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u} \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}.
%,We ignore the advective components of the material derivative and consider small
% deviations $\Upsilon' = \Upsilon - \overline{\Upsilon}, \varpi' = \varpi -
% \overline{\varpi}$ about hydrostatic equilibrium:
% \begin{subequations}\label{se:lin_homo}
%     \begin{align}
%         \bm{\nabla} \cdot \bm{u}_1 &= 0,\\
%         \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
%         \pd{\bm{u}_1}{t} + \bm{\nabla}P_1
%             + \p{1 + \Upsilon_1} g\bm{z}
%             &= 0.
%     \end{align}
% \end{subequations}
The solution in this linear regime up to undetermined amplitude $A$ is
given by \citep{drazin,sutherland0}:
\begin{equation}
    u'_z\p{x, z, t} = Ae^{z/2H}e^{i\p{k_{x}x + k_{z}z - \omega t}},
        \label{eq:k0z_sign}
\end{equation}
where $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
We take
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
the \emph{Brunt-V\"ais\"al\"a frequency} to be constant and take the
incompressible limit $c_s \to \infty$ where $c_s$ is the sound speed. Other
dynamical quantities are simply related to $u'_z$ (see
e.g.\ \cite{sutherland0}).

In the weak stratification limit $k_{z}H \gg 1$, the solution exhibits the
following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows with $z$ as $e^{z/2H}$. Thus, the
        linear approximation is always violated for sufficiently large $z$.

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \bm{c}_{ph} &=
                \p{k_{x}\bm{x} + k_{z}\bm{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\bm{x}
                - \p{k_{x}k_{z}\bm{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{ph} \cdot \bm{c}_g =
        \mathcal{O}\p{\frac{1}{\p{k_{z}H}^{2}}} \approx 0$, the usual result
        (e.g.~\cite{drazin,sutherland1}). Note again that $c_{gz} < 0$ for
        $k_{x}, k_{z} > 0$, so upwards-propagating IGW have $k_z < 0$.

    \item The time-averaged total $x$-momentum flux $F$ carried in the $\bm{z}$
        direction can be computed. Since the linear solution is separable as
        $f(x, z, t) = f(z)e^{i(k_{1x}x - \omega_1 t)}$, $x$ averaging and time
        averaging are equivalent, so we may write
        \begin{align}
            F &\equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x,
                    \label{eq:F_def},\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}'\pd{\ln \bar{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}, \omega_(k_{x}, k_{z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two plane wave solutions,above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z$), and we obtain
\begin{align}
    u_{z}'&(x, z, t) ={} \frac{Cgk_{x}^2}{\omega^2}
        \frac{1}{2ik_{z}}\frac{e^{\frac{-(k_{z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{1}{2k_{z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upwards-propagating IGW wavetrain excited deep
in the atmosphere.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\bm{u}' \cdot \bm{\nabla}$. However, since $\abs{\bm{u}'} \propto
e^{z/2H}$, any nonzero amplitude IGW will eventually violate the linearity
criterion. This regime is relevant in practice since the IGW may grow to
nonlinear amplitudes before it reaches the surface of the WD
\citep{fullerI,fullerII}. We can coarsely estimate the height at which this
happens by when the Lagrangian displacement of the perturbation $\bm{\xi}' =
\frac{\bm{u}'}{-i\omega}$ of a fluid parcel satisfies
\begin{equation}
    \abs{\bm{\xi}'} \abs{\bm{k}} \gtrsim 1.\label{eq:nl}
\end{equation}

This is in agreement with the breaking location predicted by the wave-induced
mean flow, given by \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_{x}} \equiv \overline{U} = \frac{1}{L_x}
        \int\limits u_{x}\;\mathrm{d}x = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
This formula is analogous to Stokes' drift for surface waves and can be derived
by considering the propagation of $F$ (\autoref{eq:F_def}) into a medium at
rest. Solving for $\overline{U} = \omega / k_x$ the critical layer condition
(see \autoref{ss:crit_layer} immediately below), we obtain
$\abs{\frac{u_{x}'k_{x}}{\omega}} = \abs{\xi_x k_{x}} = \abs{\xi_z k_z} \sim 1$,
equal to the nonlinearity criterion \autoref{eq:nl}. This condition predicts
where the wave's self-acceleration alone produces a critical layer.

@@

A purely horizontal shear flow $\overline{U}(z) \bm{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\overline{U}(z)\partial_x$. For a critical value $\omega - \overline{U}_c k_{x}
= 0$, the frequency of the linear wave in the fluid's frame of reference
vanishes and critical behavior is observed. In a linear
\citep{booker_bretherton} or viscous \citep{hazel} theory, the incident wave has
amplitude reflection and transmission coefficients
\begin{align}
    \mathrm{Ri} &\equiv \at{\frac{N^2}{\p{\pd{\overline{U}}{z}}^2}}_{z_c},
        \label{eq:ri_def}\\
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where we have defined local Richardson number Ri at the critical layer $z_c:
\overline{U}(z_c) = \frac{\omega}{k_{x}}$. In the $\mathrm{Ri} \gg 1$ limit,
$\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is completely absorbed to
good approximation.

When the fluid absorbs the incident wave, it absorbs the incident horizontal
momentum flux as well, which is converted into additional horizontal momentum of
the shear flow. Since the shear flow cannot exceed $\overline{U}_c$ the horizontal
phase velocity of the incident wave, the critical layer must instead propagate
downwards to accommodate the incident momentum. The total horizontal momentum of
the shear flow then obeys conservation equation
\begin{equation}
    \pd{}{t}\int\limits_0^{L_z} \rho(z) \overline{U}(z, t)\;\mathrm{d}z
        + \Delta S = 0.
\end{equation}
We define $\Delta S$ to be the change in flux across the critical layer, or
equivalently the negative of the absorbed flux. Treating $\overline{U}(z > z_c) =
\overline{U}_c, \overline{U}(z < z_c) = 0$ gives us exactly
\begin{equation}
    \rho(z_c) \overline{U}_c\pd{z_c}{t} = \Delta S(z_c(t)).\label{eq:zc_anal}
\end{equation}
$\Delta S(z_c)$ denotes the change in $S$ across $z_c$. For constant $\Delta S$
in time and $\rho \approx \bar{\rho}$, this has analytical solution
\begin{equation}
    z_c(t) = -H\ln t - H\ln \frac{H\bar{\rho}_0c_{ph, x}}{-\Delta S}
        .\label{eq:zc_sol}
\end{equation}
Note that constant $-\Delta S$ corresponds physically to the picture where
constant $S$ carried by the IGWs, given by \autoref{eq:S_lin}, is fully absorbed
at the critical layer, in which case $\Delta S = -S$.

@@

We nondimensionalize by taking $H = N = \bar{\rho}_0 = 1$ in
\autoref{se:nl_orig}.

% TODO physical vs simulation domain
Denote $L_x, L_z$ to be the physical dimensions of the simulation domain. Our
simulation domain is smaller than our full physical domain, so we adopt
reflection-suppressing boundary conditions. We use periodic boundary conditions
in the $x$ direction and damping layers (described in \autoref{ss:damping}) in
the $z$ direction to damp perturbations exiting the top and bottom of the
domain. We use a Fourier basis in both the $x$ and $z$ direction. We varied the
number of $x, z$ modes (denoted $N_x, N_z$ respectively), and used $3/2$
dealiasing \citep{boyd}.

@@

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from advective Courant-Friedrichs-Lewy (CFL)
time. Specifically, we use $\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_{z})$,
where the minimum is taken over every grid point in the domain and $\Delta
x,\Delta z$ are the grid spacings in the $x$ and $z$ directions respectively.

The physics of our simulation is fixed by four parameters: $k_{x}, \omega, C,
\nu$. We describe our choices for these parameters below:
\begin{itemize}
    \item $k_{x}$: Astrophysical IGWs in stars are generally excited by the $l
        = 2$ component of the tidal potential, for which $k_{\perp} \sim
        \frac{1}{R}$ where $R$ is the radius of the star. To best emulate this,
        we use $k_{x} = \frac{2\pi}{L_x}$ the smallest permitted wavenumber
        permitted by periodic boundary conditions.

    \item $\omega$: We choose $\omega$ by evaluating dispersion relation
        $\omega\p{k_{x}, k_{z}}$ for some desired $k_{z}$.

        Astrophysical IGWs also generally satisfy $\omega \ll N$, or
        equivalently $\frac{k_r}{k_{\perp}} \sim k_rR \gg 1$. Since $H \lesssim
        R$, we aim to study $k_{z}H \gg 1$. However, to satisfy this well is
        computationally expensive: $L_z \gg H$ is required to give waves ample
        room to grow within the simulation domain, but we want $k_{z}$ to be
        sufficiently separated from the grid spacing $\sim L_z / N_z$ that
        nonlinear effects can be seen in later simulations (see
        \autoref{ss:nl_ns}). Thus, $N_z$ sets how well we can satisfy $k_{z}H
        \gg 1$; we choose $k_{z}H = 1$.
